R Notebook

### 2024-07-29 Scripts/authoritarian_only_latex_2.Rmd
Authoritarian only, before re-write to compare non-strat + strong vs. strategic and weak

---
output:
    pdf_document:
      keep_tex: yes
linkcolor: blue
bibliography: C:/Users/colej/Documents/Research projects/bib_ch.bib
#biblio-style: "apalike"
fontsize: 12pt
link-citations: yes
toc: false
csl: C:/Users/colej/Documents/Research projects/ji_manipulation_protest/american-journal-of-political-science.csl
indent: true
abstract: "The risk of mass protest figures prominently in many models of election manipulation, as a deterrent to tampering with elections. Yet election manipulation is common outside liberal democracies, while protest is rare (and successful protest rarer still). Using data on election fraud and duration data for protest events from 647 elections from 1990 to 2012, this paper argues that the risk to incumbents has been overstated in prior work. It argues that incumbents who solve the principal-agent dilemmas associated with election fraud are likely to be highly resilient to protest. There are two main findings. First, a riskier pre-election environment for incumbents is associated with more fraud, not less. Second, when protests do occur, greater election-day fraud is associated with a faster demise for protest movements. This pessimistic finding implies that election fraud is less risky than much prior work assumes, and helps explain the long-term stability of electoral authoritarianism as a regime type."
header-includes: 
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[table]{labelformat=empty}
---

@@

Despite the significant challenge, the ruling party was able to defuse the protest movement over a period of months. Through a combination of repression and concessions [@golosov2012], the ruling party was able to reassert itself and continue a long process of sidelining the political opposition in Russia [@gelman_2015_troubled]. The post-2011 protest wave is one much-studied case of the phenomenon of electoral protest aimed at punishing electoral manipulation. Yet the Russian example raises questions for that broader phenomenon: does election fraud itself make protest more likely, or only in the context of a weak electoral showing for the ruling party? Given the ruling party's poor result and substantial electoral fraud, why was it able to dispatch the mobilized opposition with relative ease?

These questions get to the heart of a debate in the study of electoral manipulation and protest: does election fraud signify regime weakness or strength, and in what ways? The various theoretical perspectives brought to bear on the topic have been built on different assumptions, and yield different implications for authoritarian governance and democratization. For some models, election fraud is a tool of weak incumbents. As a result, incumbents avoid fraud when it might be exposed and reveal their weakness to potential protesters\textemdash an auspicious prediction for democratization and ultimately the under-pinning of 'self-enforcing' democracy [@fearon_self-enforcing_2011]. Another school of thought holds that election fraud can be an indicator of incumbent strength. This approach, which is expanded upon below, is a pessimistic account for democracy. If this model is correct, it is authoritarian politics rather than democracy that is self-enforcing\textemdash the ability to commit large-scale election fraud helps win elections, signal strength, deter challengers under most conditions, and put down protest movements if they do begin. This paper argues that this pessimistic model best fits the overall pattern of the data, and is supported by specific, novel hypothesis tests.

However, these models overlook the fact that engineering widespread fraud requires solving principal-agent problems through consolidated control over patronage resources and the criminal and judicial apparatus [@Harvey2019;@Harvey2022]. As a result, high levels of fraud are a sign of incumbent strength in both cooptation and repression (Gerschewski's third pillar). Both of these tools can be useful for defusing or dismantling an emergent protest movement. As a result, effective fraudsters should also be more effective at defeating protest movements. Consequently, in this view, there is little penalty for severe manipulation in terms of protest risk, and little deterrent effect.

Testing this proposition in more depth is difficult. If, for example, incumbents manipulate just the right amount to avoid mass protest in equilibrium\textemdash and are deterred from manipulating more by that risk\textemdash we would observe no statistical relationship between the severity of fraud and the likelihood of protest initiation. High and low levels of fraud would generally be associated with no protest. Beginning to untangle the relationship between election fraud and protest requires looking beyond standard binary measures of protest. Instead, this paper tests outcomes "off the equilibrium path," to use the language of formal modeling. Does greater risk to the incumbent in the pre-election period result in reduced election fraud? When protest movements do arise after an election, is fraud associated with weaker governments? Using data from 614 non-democratic elections across 97 countries, from 1989 to 2012, this paper finds that the answer to both questions is no. In fact, greater pre-election risk is associated with worse fraud, and more severe election fraud is associated with a *faster* demise for protest waves. This indicates that, far from being a sign of weakness, in most cases the ability to generate significant electoral fraud is an indicator of durable regime strength. This finding holds even when incumbents claim small vote-shares, suggesting that electoral fraud capacity is an independent component of regime stability.

Given these results, it is likely that the downside risk for incumbents that engage in election fraud in non-democracies has been overemphasized. Fraud itself is a sign of the robustness of some of the same tools that can be used to co-opt the opposition, constrain elites, or intensify repression during a crisis; far from being a sign of weakness, it can be a sign of resilience. This can deter protest, especially in conjunction with high claimed vote-shares for the incumbent. Even in the unlikely event that protests do break out in response to large-scale fraud, a government that has engaged in widespread fraud has many tools it may deploy to curtail them.

These results help explain several empirical patterns at once: the relative paucity of electoral protests, the rarity of their success in overturning incumbent regimes, and governments' attempts to claim implausibly high margins of victory in conjunction with high levels of fraud. At a theoretical level, they also suggest that protest risk should be de-emphasized in research on the causes of election manipulation, in favor of alternative explanations which emphasize the role of front-line actors in delivering manipulation. Finally, they suggest a pessimistic outlook for democratization by election in authoritarian regimes.

# Literature review: election manipulation and mass protest

Prior research has often considered that electoral manipulation has both costs and benefits for ruling parties. The benefits include better odds of winning the official tally [@lehoucq2002], of course, as well as any signals of strength and stability that a larger vote-share [@magaloni2006] or display of power might provide [@simpser2013;@gehlbach2015electoral]. On the other side, election manipulation is generally characterized as having both operational costs (which the incumbent may be unwilling or unable to pay [@Greene2010]) and legitimacy costs which reduce other actors' level of cooperation with the incumbent [@birch2011]. The most visible, and arguably most studied, form of such legitimacy costs is mass post-election protest against the incumbent.

The cost of protest is conceptualized as the risk that the incumbent may be unseated and/or the resource cost needed to repress or buy off the incumbent's opponents. Various theories model the decision to engage in post-election protest either as a strategic one by opposition groups [@magaloni2010;@Luo2020;@przeworski2015], or more spontaneous coordination by citizens expressing a shared grievance [@tucker2007;@Brancati2016]. In either case, the risk of protest is a classic component of models of electoral manipulation, as a restraint on incumbents who would prefer to use fraud to increase their vote-shares, all else held equal. This argument can be found in theories about the conduct of non-democratic elections, as well as in the literature on self-enforcing democracy [@fearon_self-enforcing_2011;@przeworskiSelfenforcingDemocracy2006]. In the latter literature, generally free and fair elections are thought to be a self-enforcing equilibrium through the threat of mass protest by wronged groups if the incumbent government rigs the results. While the focus here is on elections in authoritarian cases, some implications for self-enforcing democracy are discussed at the close of the paper.

<!-- Broadly, prior work has considered two major explanations for the severity of manipulated elections: the costs and benefits of manipulation to the ruling party and government (a demand-side framework) and the costs and benefits for the front-line actors who do the work of manipulating elections (a supply-side framework). The risk of costly protest is the central, though not sole^[In some studies, like [@birch2011], the costs of manipulation are attributed to a more diffuse loss of legitimacy; such costs *may* manifest as protest, but may also manifest as a reduction in general cooperation with the regime. However, it is worth noting the argument of [@gehlbach2015electoral] that election manipulation can *induce* cooperative behavior from some agents.], factor which affects the demand-side cost of manipulation in most models. -->

<!-- In the major formulations of this argument, such as Fearon  [-@fearon_self-enforcing_2011] and Przeworski [-@przeworskiSelfenforcingDemocracy2006], 'democracy' is understood in procedural terms. For Przeworski, democracy requires that "authorization to rule is derived from elections" which designate winners and losers according to a set of rules [@przeworskiSelfenforcingDemocracy2006, p. 312]. Fearon adopts a similar definition for electoral democracy [@fearon_self-enforcing_2011, p. 1662]. Such arrangements are not limited to liberal democracies, and in fact constitute the large majority of regimes today. Many previous works argue that the risk of protest can mitigate election manipulation in regimes where elections are held\textemdash including established democracies [@millerRepublicIfYou2021], electoral authoritarian regimes [@Egorov2021], and any regime where elections are held but may be manipulated [@Little2015;@Luo2020]. In most such models, protests are successful in overthrowing the government with some probability, but impose a cost on both the protesters and the government regardless of success (e.g. [@Luo2020]); the logic thus extends even to highly repressive and entrenched regimes. Nevertheless, this body of work has yet to grapple with the empirical finding that election integrity is relatively poor across these regimes outside of a narrow club of 'consolidated' democracies.  -->

Much of the work on this subject has been done in the formal modeling tradition. In game-theoretic research where the risk of protest is shown to deter fraud, the election is usually conceived as a signaling game. In such models, incumbents are characterized as 'weak' or 'strong' based on their support in the electorate; weak incumbents may need to resort to fraud in order to secure re-election. As a result, the opposition perceives election fraud as a reliable signal of incumbent weakness, and will stage protests in order to push out the weak incumbent. For example, the model developed by Little, Tucker, and LaGatta [-@Little2015] predicts that "Just as stronger election results for the incumbent lead to less protest because they indicate that she is popular, reports of more fraud lead to more protest because they indicate that the incumbent is, for a given election result, less popular" (p. 1153). In equilibrium, this can lead weaker incumbents to step down rather than face protest due to massive fraud\textemdash a pro-democracy outcome. Similar results are developed by Chernykh and Svolik [-@chernykh2015third], Luo and Rozenas [-@Luo2020] and [-@Luo2017], and Magaloni [-@magaloni2010]. These are normatively optimistic models upon which a sizable literature has been founded: fraud can be kept in check, and ultimately reduced, through the development institutions that reveal information about election integrity: election monitoring [@Hyde2014;@daxecker2012], more independent courts [@chernykh2015third], media activity [@naiFourthEstate], and civil society coordination [@birch2017getting].

Another school of thought holds that electoral protest is driven by citizens' grievances rather than by a strategic, organized opposition [@Brancati2016]. In this view, election manipulation can act as a shared grievance among individuals who prefer to see the ruling party lose\textemdash especially when it appears that fraud may have been decisive [@kuntz_more_2009]. Belief in a stolen election is a particularly motivating kind of grievance, because it acts to focus all the diverse personal grievances individuals may have against the regime into one moment of opportunity for action [@tucker2007], possibly initiating a protest cascade [@Kuran1991]. Just as above, these models tend to predict greater protest risk when margins of victory are low, suggesting a stolen election, and/or grievances are high in the presence of fraud.

@@

This implication connects to a second class of formal models, which hold that fraud operates as a distortion of the signal of the incumbent's true strength. In this type of model, such as those developed by Little [-@littleFraudMonitoringNoncompetitive2015] or Egorov and Sonin [-@Egorov2021], fraud may be used by incumbents to bluff their opponents into being more supportive than they would be if they knew the incumbent's true popularity. However, greater transparency makes such bluffs more costly. The typical result in these models is that protest risk can *partially* deter fraud. In equilibrium, the incumbent will generally choose the level of fraud anticipated by the opposition and no more\textemdash but also no less.

Finally, there is a third school of thought, which holds that election manipulation may indicate regime strength, especially if manipulation is blatant or produces large margins of victory [@simpser2013]. In the next section, I expand on this school of thought in more detail.

There is a general, though not complete, consensus in empirical work that election fraud is associated with increased risk of protest. Several cross-national studies have found links between election manipulation and binary measures of protest [@Brancati2016;@Hyde2014;@Harvey2020;@Rod2019]. A study of African elections from 1997 to 2009 shows that election fraud is associated with an increased number of post-election conflicts [@daxecker2012]. However, Hafner-Burton et al [-@Hafner-Burton2016] find that the positive association between election fraud and post-election protest falls short of conventional statistical significance. Studies of one archetypal example, the 2011 election and protest cycle in Russia, have found a positive relationship between fraud and protest by some measures [@Lankina2017], and no relationship by others [@ananyevDictatorsSignalStrength2022]. Taking an alternative approach, Rozenas [-@Rozenas2016b] finds that governments facing pre-election protests manipulate less severely; however, the dependent variables for that study are largely defined by the structural conditions of the election (i.e. whether opponents are allowed to run), rather than by the election-day manipulation that usually figures in theoretical work. This result is consistent with the finding that parties are more likely to protest elections after ex ante changes to the rules [@chernykh2014].

While these works have identified a connection between fraud and protest, in general, they cannot distinguish between the competing assumptions of the two theories discussed here, or the third possible approach discussed in below. For instance, all three approaches predict that protest should be most likely when evidence of fraud is high and the incumbent's vote-share is low (or, relatedly, grievances are high). Such a scenario could occur due to 1) strategic protest against an incumbent revealed to be weak through reliance on fraud, 2) mass-led protest against a 'stolen' election, or 3) risky strategic protest against a low-popularity incumbent despite institutional reservoirs of strength. The implications of all three scenarios are substantively distinct, however. To untangle them, it is necessary to test the assumptions underlying the different models. In particular, do post-election protest movements encounter weak incumbents as signaling models predict?  If not, the deterrence model is undermined.

<!--  One critique of this view, although still within the information/strategic framework, comes from Simpser's [-@Simpser2011] pathbreaking book. Simpser argues that when wanipulation efforts are excessive---well beyond what is needed simply to win---they demonstrate the organizational capacity, resources, and manpower of the ruling party [@Simpser2013]. Large-scale manipulation is, thus, a signal of strength rather than weakness, and can deter protest. By contrast, partial manipulation efforts that fail to secure a large majority vote-share signals the government lacks both popular support and mobilizational capacity, making opposition protest more likely. -->

@@

As discussed above, one view of election fraud and protest views fraud as a tool of weak incumbents. Another view, associated with Simpser [-@simpser2013], holds that election fraud demonstrates incumbent strength. In Simpser's original account, we should expect electoral manipulation to signal strength when it is readily observable or excessive, producing large margins of victory. A strong government may manipulate heavily in order to produce a very large majority [@magaloni2006], to induce regime insiders to remain cooperative [@gehlbach2015electoral], or because of a bandwagon effect among front-line agents [@rundlett2016deliver]. This argument is in keeping with Seeberg's [-@Seeberg2019] finding that higher-capacity states are weakly associated with more election fraud.

I build on Simpser's logic, and on subsequent research, by arguing that that extensive election fraud itself requires the use of resources that are sources of incumbent strength. Simpser's model is concerned with whether or or not other actors can perceive the regime's signal of strength; hence the focus on excessive or blatant manipulation. I argue here that even when large-scale election fraud is marginal (i.e. produces narrow margins of victory), it can indicate regime strength independent of popularity. This argument  draws on the recent strand of research on election manipulation that emphasizes the principal-agent problems inherent in such campaigns. Political principals require the cooperation of large numbers of front-line agents in order to generate extensive election fraud, agents who may shirk if their perceived interests are not aligned with the incumbent [@rundlett2016deliver].

To the extent that election results in non-democracies send a signal, it is composed of a mix of three factors [@Pop-Eleches2015]: the incumbent's genuine popularity, its ability to mobilize otherwise reluctant voters (through patronage, clientelism, and coercion), and its ability to control the bureaucratic apparatus (including the criminal and judicial systems) to generate fraud. Large-scale election fraud requires that the incumbent have consolidated control over patronage resources in society; this makes cooperation with the incumbent attractive for front-line agents, who are less willing to risk losing access to the incumbent's network [@Harvey2019]. However, such efforts also require that incumbents be able to shield front-line agents from negative consequences if their illegal efforts are publicly exposed [@Harvey2022]. In other words, the incumbent must have sufficient control over the police and the courts, such that ordinary individuals expect little punishment for committing criminal election fraud on their behalf. Each of these capacities is required to commit large-scale fraud, and both are useful for disarming protest movements. To summarize, the more extensive the election-fraud effort, the more successfully the incumbent has solved the associated principal-agent problems, and the more likely it will be able to use those tools to quash protest.

The initial emergence of protest triggers a response by the government and ruling party that can take the form of repression, concessions, or toleration [@tillyMobilizationRevolution1978;@lipskyProtest1968;@Hummel2019;@raslerConcessionsRepression1996;@Bishara2015;@Frantz2014]. Concessions may be aimed at protesting parties themselves [@reuter2015;@mooreRepressionDissentSubstitution2000;@Hafner-Burton2016;@Piven1979] or the more moderate public, in order to keep them on the sidelines of the conflict [@Hummel2019]. In either case, large-scale election fraud indicates control over considerable patronage resources, which in turn indicates the capacity to engage in such payments (or withhold them from clients in exchange for passive support [@Rosenfeld2017]).

>>>>

---
output:
    bookdown::pdf_document2:
    keep_tex: true
linkcolor: blue
bibliography: C:/Users/colej/Documents/Research projects/bib_ch.bib
#biblio-style: "apalike"
fontsize: 12pt
link-citations: yes
toc: false
csl: C:/Users/colej/Documents/Research projects/ji_manipulation_protest/american-journal-of-political-science.csl
indent: true
abstract: "The risk of mass protest figures prominently in many models of election manipulation, as a deterrent to tampering with elections. Yet election manipulation is common outside liberal democracies, while protest is rare (and successful protest rarer still). Using data on election fraud and duration data for protest events from 647 elections from 1990 to 2012, this paper argues that the risk to incumbents has been overstated in prior work. It argues that incumbents who solve the principal-agent dilemmas associated with election fraud are likely to be highly resilient to protest. There are two main findings. First, a riskier pre-election environment for incumbents is associated with more fraud, not less. Second, when protests do occur, greater election-day fraud is associated with a faster demise for protest movements. This pessimistic finding implies that election fraud is less risky than much prior work assumes, and helps explain the long-term stability of electoral authoritarianism as a regime type."
header-includes: 
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[table]{labelformat=empty}
  - \usepackage{multirow}
  - \usepackage{lscape}
  - \usepackage{pdflscape}

---

@@

Despite the significant challenge, the ruling party was able to defuse the protest movement over a period of months. Through a combination of repression and concessions [@golosov2012], the ruling party was able to reassert itself and continue a long process of sidelining the political opposition in Russia [@gelman_2015_troubled]. The post-2011 protest wave is one much-studied case of the phenomenon of electoral protest aimed at punishing electoral manipulation. Yet the Russian example highlights questions that get to the heart of a debate in the study of electoral manipulation and protest. When opposition movements protest manipulated election results, are they doing so because the incumbent\textemdash by relying on manufactured votes\textemdash has signaled its own weakness? This view is adopted by many researchers, especially in the formal modelling tradition. Or, does election manipulation indicate sources of strength for incumbents (independent of genuine popularity), as others have held?

The stakes of this debate are significant. The former view is an optimistic one; incumbents avoid fraud when it might be exposed and reveal their weakness to strategic protesters\textemdash an auspicious prediction for democratization and ultimately the under-pinning of 'self-enforcing' democracy [@fearon_self-enforcing_2011]. If the latter view holds, electoral protest may mainly occur non-strategically, and would typically fail to deter election manipulation as a result. Mass protest would, in this case, be a fragile guardrail against autocratization. This paper argues that this pessimistic model best fits the overall pattern of the data, and is supported by specific, novel hypothesis tests.

This paper contributes to this debate by drawing out the implications of three strands of the literature on election manipulation and protest. In short: there are two views on the nature of electoral protest which make the same prediction. If opposition groups strategically initiate protest when they believe the incumbent to be weak, they will be most likely to do so when the election results show a low vote-share for the incumbent combined with high levels of manipulation\textemdash a combination of factors which reveals the incumbent's very low genuine popularity. However, if protest movements are more organic and non-strategic, reflecting citizen grievances, we should expect protest initiation in the same circumstances. Here the combination of low incumbent vote-share and high manipulation indicates severe citizen grievances (low natural popularity) which are galvanized by the shared grievance of a 'stolen' election. Since both theories make the same prediction about protest initiation, such tests\textemdash which have been the dominant empirical measure in the literature\textemdash do not reveal whether protest is strategic and manipulation indicates weakness or manipulation indicates sources of strength and anti-manipulation protest is non-strategic.

To unravel this question, I make use of data on the duration of post-election protests. Shorter-lived protest movements that end with the incumbent still in power indicate a government that is relatively strong compared to the protesting opposition, while longer-lived protest movements indicate a more even balance of power between the two. The reverse is also true: shorter-lived protest movements that end in opposition victory indicate a weaker government, while governments that hold on longer before being defeated are relatively stronger. Using data from 614 non-democratic elections across 97 countries, from 1989 to 2012, this paper finds that 1) protest initiation is most likely when manipulation is severe but incumbent vote-share is low, as both theories predict, and 2) that governments that engage in more manipulation are likely to put down protests more quickly than governments that produce little manipulation, especially when incumbents claim small vote-shares.

This pattern in the results is consistent with the argument that, rather than being a sign of weakness, the ability to generate significant electoral fraud is an indicator of durable regime strength. It is also consistent with the argument that much post-election protest in non-democracies is non-strategic; i.e., the actions of a weak opposition against a strong incumbent. Given these results, it is likely that the downside risk for incumbents that engage in election fraud in non-democracies has been overemphasized. This interpretation helps explain several empirical patterns at once: the relative paucity of electoral protests, the rarity of their success in overturning incumbent regimes, and governments' attempts to claim implausibly high margins of victory in conjunction with high levels of fraud. At a theoretical level, they also suggest that protest risk should be de-emphasized in research on the causes of election manipulation, in favor of alternative explanations which emphasize the role of front-line actors in delivering manipulation. Finally, they suggest a pessimistic outlook for democratization by election in authoritarian regimes.

# Literature review: three views of election manipulation and mass protest

Three understandings of the relationship between election manipulation and post-election protest have developed in the literature. Each view is discussed in detail below, after a brief initial summary here. A first view assumes that incumbents and opposition groups are strategic actors who who evaluate the election period for signs of each other's relative likelihood of winning a power struggle. Researchers who model election protest this way generally treat the use of election manipulation as a sign of the incumbent's unpopularity, and therefore weakness. Protest occurs after fraudulent elections in response to this perceived weakness; this possibility can serve to deter incumbents from engaging in manipulation under some circumstances. This is a normatively optimistic theory.

A second worldview argues that election manipulation can indicate a well-resourced, well-organized, powerful incumbent (independent of its popularity). Under this view, extensive election manipulation can be a show of strength, which induces cooperation with the incumbent from other societal actors. Since this view holds that more manipulation can further entrench the incumbent, it is a normatively pessimistic view.

Finally, others have argued that electoral protest occurs unrelated to strategic concerns about the relative strength of the incumbent and the opposition. In these models, electoral protest occurs because manipulation galvanizes citizens who hold diverse grievances, because economic conditions are poor, or simply because holding an election creates a moment of opportunity. The central feature of these models is that protest decisions are generally made without consideration for the likelihood that protests will overthrow the incumbent. As will be discussed in depth below, these different assumptions yield different predictions for when electoral protest is more likely to be initiated, and how long it will last against an incumbent that refuses to give up power. Table 1 presents these predictions for ease of reference.

```{=latex}
% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
% \usepackage{multirow}
\begin{table}[]
\begin{tabular}{@{}llll@{}}
\toprule
                                                                                                              & Manipulation signals: & \multicolumn{1}{c}{Strategic protest}                                                           & \multicolumn{1}{c}{Non-strategic protest}                                                  \\ \midrule
\multicolumn{1}{l|}{\multirow{2}{*}{Initiation most likely}}                                                  & Weakness              & \multicolumn{1}{l|}{\begin{tabular}[c]{@{}l@{}}Low margin + \\ high manipulation\end{tabular}}  & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Low margin + \\ high manipulation\end{tabular}} \\ \cmidrule(lr){2-3}
\multicolumn{1}{l|}{}                                                                                         & Strength              & \multicolumn{1}{l|}{\begin{tabular}[c]{@{}l@{}}Low margin + \\ low manipulation\end{tabular}}   &                                                                                            \\ \cmidrule(l){2-4} 
\multicolumn{1}{l|}{\multirow{2}{*}{Initiation least likely}}                                                 & Weakness              & \multicolumn{1}{l|}{\begin{tabular}[c]{@{}l@{}}High margin + \\ low manipulation\end{tabular}}  & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}High margin +\\ low manipulation\end{tabular}}  \\ \cmidrule(lr){2-3}
\multicolumn{1}{l|}{}                                                                                         & Strength              & \multicolumn{1}{l|}{\begin{tabular}[c]{@{}l@{}}High margin + \\ high manipulation\end{tabular}} &                                                                                            \\ \cmidrule(l){2-4} 
\multicolumn{1}{l|}{\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Duration when \\ incumbent wins\end{tabular}}} & Weakness              & \multicolumn{2}{c}{More manipulation: Longer protest}                                                                                                                                        \\
\multicolumn{1}{l|}{}                                                                                         & Strength              & \multicolumn{2}{c}{More manipulation: Shorter protest}                                                                                                                                       \\ \cmidrule(l){2-4} 
\end{tabular}
\caption{Predictions for protest initiation and duration under varying theoretical assumptions}
\label{tab:predictions}
\end{table}
```

In classic models of election manipulation, manipulation is a short-term tool used instrumentally to secure victory in the election [@lehoucq2002]. Much of the work in this tradition has been elaborated through formal models. Though there are, of course, many variations, in several such models incumbents are characterized as 'weak' or 'strong' based on their true support in the electorate; weak incumbents may need to resort to fraud in order to secure re-election. As a result, the opposition perceives election fraud as a reliable signal of incumbent weakness, and will stage protests in order to push out the weak incumbent. For example, the model developed by Little, Tucker, and LaGatta [-@Little2015] predicts that "Just as stronger election results for the incumbent lead to less protest because they indicate that she is popular, reports of more fraud lead to more protest because they indicate that the incumbent is, for a given election result, less popular" (p. 1153). In equilibrium, this can lead weaker incumbents to step down rather than face protest due to massive fraud\textemdash a pro-democracy outcome. Similar results are developed by Chernykh and Svolik [-@chernykh2015third], Luo and Rozenas [-@Luo2020] and [-@Luo2017], and Magaloni [-@magaloni2010].

A related class of formal models holds that fraud operates as a distortion of the signal of the incumbent's true strength. In this type of model, such as those developed by Little [-@littleFraudMonitoringNoncompetitive2015] or Egorov and Sonin [-@Egorov2021], fraud may be used by incumbents to bluff their opponents into being more supportive than they would be if they knew the incumbent's true popularity. However, greater transparency makes such bluffs more costly. The typical result in these models is that protest risk can *partially* deter fraud. In equilibrium, the incumbent will generally choose the level of fraud anticipated by the opposition and no more\textemdash but also no less.

Fundamentally, the set of models which assume that protest is a centralized, strategic decision by opposition groups predict that the highest risk of protest should obtain when incumbents use massive fraud to generate a low incumbent vote-share, indicating that the incumbent's true popularity is very low. Conversely, they predict that protest should be least likely when the incumbent's vote-share is high and manipulation low, indicating a high level of true popularity that deters opposition protest.

The models described above conceive of opposition protest to election manipulation as a strategic action undertaken because the use of more manipulation increases the likelihood that the opposition will have support in the streets (due to the incumbent's low popularity). These are normatively optimistic models upon which a sizable literature has been founded: fraud can be kept in check, and ultimately reduced, through the development institutions that reveal information about election integrity: election monitoring [@Hyde2014;@daxecker2012], more independent courts [@chernykh2015third], media activity [@naiFourthEstate], and civil society coordination [@birch2017getting].

A second set of models shares the assumption that protest is a strategic action motivated by how confident opposition groups are they will succeed in a conflict with the incumbent. It differs, however, in its conceptualization of the information revealed by election manipulation. Most persuasively articulated by Simpser [-@simpser2013], this view holds that election manipulation demonstrates incumbent strength. As he writes, an incumbent that demonstrates the ability to engage in widespread manipulation "is also likely to
have the resources, capacities, and inclinations to overcome or punish opponents, reward supporters, and circumvent the law" [@simpser2013, p. 6]. A strong incumbent (in the language used by the first set of models, i.e. one that is relatively popular) may nonetheless manipulate heavily in order to produce a very large majority [@magaloni2006], to induce regime insiders to remain cooperative [@gehlbach2015electoral], or because of a bandwagon effect among front-line agents [@rundlett2016deliver]. By demonstrating otherwise difficult-to-observe capabilities, large-scale election manipulation can solve a coordination problem for the incumbent, securing the cooperation of elites [@simpser2013]. This can, in turn, fix expectations that the incumbent will continue to rule in the future, further enhancing its staying power [@hale_patronal_2014]. This argument is in keeping with Seeberg's [-@Seeberg2019] finding that higher-capacity states are associated with more election fraud.

Since these views hold that large-scale election manipulation can reveal information about sources of incumbent strength other than true popularity, and can have indirect, positive effects on incumbent stability, they are normatively pessimistic. They predict that opposition protest will be most likely when incumbent vote-share is low and manipulation is low, indicating that the incumbent lacks both genuine popularity and the resources and capacities needed to manufacture votes. By contrast, protest should be least likely to occur, under this framework, when incumbent vote-share and manipulation are both high.

Finally, a third school of thought holds that decisions to engage in electoral protest are non-centralized, driven by citizens' grievances rather than by a strategic, organized opposition [@Brancati2016]. In this view, election manipulation can act as a shared grievance among individuals who prefer to see the ruling party lose\textemdash especially when it appears that fraud may have been decisive [@kuntz_more_2009]. Belief in a stolen election is a particularly motivating kind of grievance, because it acts to focus all the diverse personal grievances individuals may have against the regime into one moment of opportunity for action [@tucker2007], possibly initiating a protest cascade [@Kuran1991]. These models thus tend to predict that protests are most likely when margins of victory are low and election manipulation high, suggesting both a high level of grievance in society and a "stolen" election. This framework predicts protest is least likely when incumbent vote-share is high and manipulation low (i.e. low grievances and low likelihood of a stolen election).

@@

Table 1 summarizes these predictions. As can be seen in the table, if we focus only on protest occurrence as a dependent variable, we cannot distinguish between two possible data-generating processes: one in which electoral protest is strategic and manipulation signals weakness, and one in which electoral protest is largely un-centralized but manipulation indicates regime strengths. Which of these data-generating processes predominates is of significant theoretical importance of understanding authoritarianism and democratization, since the latter view suggests much dimmer prospects for democratization than the former.

This ambiguity may help account for the inconsistent relationship between election manipulation and protest occurrence found in the literature. Several cross-national studies have found links between election manipulation and binary measures of protest [@Brancati2016;@Harvey2020]. For instance, Hyde and Marinov [-@Hyde2014] use the NELDA dataset to show that protest is more likely when international observers "seriously questioned the winner of the election or the legitimacy of the process," across 758 elections (p. 343). While that study did not include data on incumbents' vote-shares, such a measure of manipulation comes closest to the 'stolen' election framework. A study of African elections from 1997 to 2009 shows that election fraud is associated with an increased number of post-election conflicts [@daxecker2012]. Similar results are found by Rød [-@Rod2019], using a different measure of manipulation.  Using an alternative measure of manipulation from the NELDA dataset, however, Hafner-Burton et al [-@Hafner-Burton2016] find that the association between election fraud and post-election protest falls short of conventional statistical significance. Studies of one archetypal example, the 2011 election and protest cycle in Russia, have found a positive relationship between fraud and protest by some measures [@Lankina2017], and no relationship by others [@ananyevDictatorsSignalStrength2022]. As noted earlier, Simpser finds a negative relationship between the severity of election manipulation and the likelihood of protest [@simpser2013] and a positive relationship between 'excessive manipulation' (i.e. extensive manipulation and a large incumbent vote-share) and regime tenure.

Since the empirical record is mixed, and two theories make similar predictions with regard to protest initiaiton, it is necessary to generate new hypotheses that can distinguish between the three models of election manipulation and protest that have been advanced. In particular, do post-election protest movements encounter weak incumbents as signaling models predict? If not, the deterrence model is undermined.

@@

Along with the political benefits it may provide, election manipulation is generally characterized as having both operational costs (which the incumbent may be unwilling or unable to pay [@Greene2010]) and legitimacy costs which reduce other actors' level of cooperation with the incumbent [@birch2011]. The most visible, and arguably most studied, form of such legitimacy costs is mass post-election protest against the incumbent. Here, I follow the broad consensus in the literature that authoritarian incumbents find protest undesirable, since they entail some risk of incumbent defeat and may require costly repression or concessions. I assume, as a result, that incumbents seek to demobilize protest; protest movements that survive longer indicate the incumbent is relatively weaker compared to the opposition. By contrast, when protest movements meet a quick demise, the incumbent can be considered relatively strong.^[In both cases, this assume that incumbent survives the protest movement, by far the modal outcome. When the opposition wins, an outcome considered in some of the empirical models below, the assumption is reversed: a quicker opposition victory indicates a weaker incumbent.] How might election manipulation be connected to the incumbent's ability to quickly defeat a protest movement?

Integrating Simpser's argument with recent research on the principal-agent dynamics inherent to election manipulation provides insights into this question. Where Simpser argues that manipulation should be blatant and excessive (that is, producing a large margin of victory) to send signals of strength, the principal-agent framework for understanding manipulation suggests otherwise. This research trend emphasizes the degree to which political principals require the cooperation of large numbers of front-line agents in order to generate extensive election fraud, agents who may shirk if their perceived interests are not aligned with the incumbent [@rundlett2016deliver]. Factors that minimize that principal-agent problem have been shown to facilitate greater election manipulation. These same factors\textemdash the incumbent's local popularity [@rundlett2016deliver], the incumbent's consolidated control over patronage resources [@Harvey2019@;@hale_patronal_2014], and low risk of punishment for pro-regime lawbreaking [@Harvey2022]\textemdash are also independently useful for suppressing post-election protest. Consequently, in an incumbent is able to manufacture a large numbers of votes, this is a signal of resource capacity, control over bureaucratic institutions, and perhaps deep wells of popularity in parts of the country. These signals remain independent sources of regime durability, even if the incumbent's margin of victory is low.

To the extent that election results in non-democracies send a signal, it is composed of a mix of three factors [@Pop-Eleches2015]: the incumbent's genuine popularity, its ability to mobilize otherwise reluctant voters (through patronage, clientelism, and coercion), and its ability to control the bureaucratic apparatus (including the criminal and judicial systems) to generate fraudulent votes. Large-scale election manipulation is enabled by consolidated control over patronage resources in society; this makes cooperation with the incumbent attractive for front-line agents, who are less willing to risk losing access to the incumbent's network [@Harvey2019]. However, such efforts also require that incumbents be able to shield front-line agents from negative consequences if their illegal efforts are publicly exposed [@Harvey2022]. In other words, the incumbent must have sufficient control over the police, the courts, and the media, such that ordinary individuals expect that punishment for illegal election manipulation is unlikely. At the same time, these capacities are useful for disarming protest movements, as discussed below.

The initial emergence of protest triggers a response by the government and ruling party that can take the form of repression, concessions, or toleration [@tillyMobilizationRevolution1978;@lipskyProtest1968;@Hummel2019;@raslerConcessionsRepression1996;@Bishara2015;@Frantz2014]. Concessions may be aimed at protesting parties themselves [@reuter2015;@mooreRepressionDissentSubstitution2000;@Hafner-Burton2016;@Piven1979] or the more moderate public, in order to keep them on the sidelines of the conflict [@Hummel2019]. In either case, large-scale election manipulation indicates control over considerable patronage resources [@@Harvey2019@;@hale_patronal_2014;@greene_why_2007], which in turn indicates the capacity to engage in such payments (or withhold them from clients in exchange for passive support [@Rosenfeld2017]).
